MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments

### 2021-10-07 main/MSstatsPTM_Main.tex
Updated main per Olgas comments, added sample size analysis

\author[1]{Devon~Kohler}
\author[2]{Tsung-Heng~Tsai}
\author[1]{Ting~Huang}
\author[4]{Erik~Verschueren}
\author[3]{Trent~Hinkle}
\author[3]{Meena~Choi}
\author[1]{Olga~Vitek}
\affil[1]{Khoury College of Computer Science, Northeastern University, Boston, MA, USA}
\affil[2]{Kent State University, Kent, OH, USA}
\affil[3]{MPL, Genentech, South San Francisco, CA, USA}
\affil[4]{Galapagos, Mechelen, Antwerp, Belgium}

@@

The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs). Experiments targeting PTMs face several specific challenges. These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression. Due to these challenges, a robust approach to estimate relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment. We propose a general statistical model and workflow that is both reproducible and comprehensive. The method measures modified and unmodified peptide abundance by summarizing intensities through Tukey’s median polish method. Then a model based on the family of linear mixed-effects models is fit. This model is automatically adjusted to the specific experimental design. Finally, the PTM abundances are adjusted to remove variance from changes in the overall protein. We implement this model in the free and open-source R package $MSstatsPTM$.

@@

The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches.\cite{Deribe} \cite{Cohen}  Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS).\cite{Kall:2011ub} \cite{Roepstorff} Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions.\cite{Mann} Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons. First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling. Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide). Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward.\cite{Olsen:2013} Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.

@@

The proposed approach was implemented using the R coding language and evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations. The approach was compared against the commonly applied t-test and Limma methods.\cite{Ritchie_15a} The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate. Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT.\cite{Choi:2014} \cite{Huang:2020}

@@

This section describes the datasets and experiments used to evaluate the proposed method. Further details can be seen in Supplemental Sections \ref{supp-sec:sim} and \ref{supp-sec:bio_investigations}

@@

The proposed statistical approach methods were evaluated using two computer simulations, with known ground truth. Specifically, their properties under adjustment with respect to protein abundance were evaluated. Two simulations were ran, mimicking a label free experiment, one including a high number of features and no missing values, and one with few modified features and including missing values.

Differential intensity levels of modified peptides may be due to changes in modification, change in protein abundance, or both. The proposed approach adjusts the abundance with respect to unmodified peptides by combining the inference of modified and unmodified peptide abundances. Alternatively, two-sample t-test or Limma that takes as input the ratio between modified and unmodified peptide intensities (difference on log scale) is commonly applied for the same purpose. In real experiments, multiple inter-related conditions are often compared together. Whereas t-test uses measurements from the two conditions being compared, the proposed approach and Limma leverage measurements in all conditions for the inference of underlying abundance. We evaluated the impact with such adjustment by computer simulation, considering the following factors: with/without protein-level changes, effect size, number of replicates and number of conditions. Further details of how the simulations were generated can be found in Supplementary Sec. \ref{supp-sec:sim}.

\paragraph*{Computer simulation 1 - label free}

@@

\paragraph*{Computer simulation 2 - missing values and low replicates}

@@

\subsection*{SpikeIn benchmark - KGG enriched \todo{label-free?}}

@@

\subsection*{Human-1mix-TMT - ubiquitination}

The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD).\cite{LUCHETTI2021} Multiplex proteomics was used to quantify the abundance of total protein and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.

\subsection*{Mouse-2mix-TMT - phosphorylation}

Here an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri). \cite{Maculins} Multiplex proteomics was used to quantify the abundance of total protein, and phosphorylation in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours). Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications. Two mixtures using 11-plex were ran over the six conditions. The six conditions were split between 11 channels leading to the experimental design being unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.

\subsection*{Human label free quantification - no global profiling run}

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with Limma were also proposed.\cite{Zhu} While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while this approach can be applied to experiments targeting PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ).\cite{Breitwieser:2013} Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.

Further details \todo{clarify which details}  on existing methods can be seen in Supplementary \ref{supp-sec:ttest}.

@@

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. In addition mixture may be included for features acquired via tandem mass tag (TMT) methods. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance.\cite{Choi:2014} In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site.\cite{Tukey:1977} After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models, taking into account the specific experimental design \todo{weird reference?} \ref{fig:data-structure}.\cite{Bolker2009} \cite{Faraway:2006} Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE. To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method. \cite{satterthwaite:1946} Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$.\cite{Benjamini:1995} More details are provided in Supplementary Sec \ref{supp-sec:test}

@@

$MSstatsPTM$ gives the ability to impute missing feature intensities if desired. When values are imputed, it is assumed they are missing for reasons of low abundance. Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model.\cite{Tukey:1977} In order to impute a feature's missing values, the feature must be present in at least one MS Run. If the feature is not present across all MS Runs, the value will be left missing. Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers. Missing value imputation is done separately for both the PTM and global protein datasets.

@@

The statistical modeling approaches discussed above can also be extended to Tandem Mass Tag (TMT) labeling methods. TMT experiments introduce an additional source of variation in the form of different mixtures. To account for mixture variance, first feature intensities are summarized again using Tukey's Median Polish and missing values are imputed, if desired, then a new linear mixed effects model with an added term for the mixture is included. The workflow and statisical model described generally follow the methods used for modeling proteins in $MSstatsTMT$. \cite{Huang:2020} These methods are repeated to quantify both modified and unmodified peptide abundance. Once both modified and unmodified peptides are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods described previously.

@@

@@

\todo{In the caption of each figure, could you clarify which dataset the figure refers to?}

@@

\subsubsection*{Computer simulation 1 - label free}

Here we see that not accounting for changes in unmodified peptide level results in a high false discovery rate, seen in Figure \ref{fig:fdr_all_models}. When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR. Recall and accuracy of the proposed approach and Limma performed similarly, with Limma performing slightly better with low replicates, while t-test lagged behind. Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power with increased number of conditions. Further results are provided in Supplementary Figures \ref{supp-fig:sim1_fpr} to \ref{supp-fig:sim1_recall}.

\subsubsection*{Computer simulation 2 - missing values and low replicates}

In this simulation, changes in unmodified peptide level still needed to be accounted for to control the FDR. Once controlling for changes in unmodified peptide, it is clear that the proposed method performed strongest across all methods, as seen in Figure \ref{fig:sim_accuracy}. The proposed method well calibrated model accuracy and the recall rate, even when the number of replicates were low. Finally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change. In comparison, Limma and t-test showed much wider distributions. This fold change comparison can be seen in Figure \ref{fig:sim_boxplot}. Further results are provided in Supplementary Figures \ref{supp-fig:sim2_recall} to \ref{supp-fig:fc_boxplot}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{SpikeIn benchmark - KGG enriched}

@@

Further results of this experiment are provided in Supplementary Sec. \ref{supp-sec:benchmark}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Human-1mix-TMT - ubiquitination}

@@

Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:ipah}.

\subsubsection*{Mouse-2mix-TMT - phosphorylation}

@@

Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.  \todo{Maybe clarify what kind of results}

\subsubsection*{Human label free quantification - no global profiling run}

@@

Further details and results of this experiment are provided in Supplementary Sec. \ref{supp-sec:usp30}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code availability}

\todo{This belongs to the 'implementation' section}

{ \bf Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html

\noindent{\bf{Github:} https://github.com/Vitek-Lab/MSstatsPTM}

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/fig1.png}
\caption{\todo{Meena?: Write about how MS for PTMs is run?}}
\label{fig:workflow}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{All the considered methods in simulation 1 correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
}\label{fig:fdr_all_models}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{The advantage of using the proposed approach is apparent when looking at simulation 2, which includes limited observations and the presence of missing values. In overall accuracy, the proposed method performs stronger than Limma and t-test in nearly every model. Even at lower replicates the proposed method still outperformed Limma. The lowest performaning method was $t$-test. Limma shows comparable performance to the proposed method in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
\label{fig:sim_accuracy}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_FC_boxplot.png}
\caption{In the second simulation with missing values and low PTM features all methods correctly estimated the fold change with a median log change of .75. The proposed method in this simulation had a visibly tighter distribution around the median when compared to Limma and $t$-test. While the mean estimated fold changes were similar, the proposed method exhibited a stronger performance in correctly estimating the fold change for all peptides.}
\label{fig:sim_boxplot}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/spike_in_msstatsptm_volcano.png}
\caption{Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment. After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation. Additionally the background grey colored peptides showed many false positives before adjustment. After adjustment the false positives were decreased considerably.
}\label{fig:spikein_prop_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
\caption{Limma method used to model the spike in experiment. Using Limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level. However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue. In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer after.
}\label{fig:spikein_limma_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
\caption{The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment. More PTMs became insignificant after adjustment then became significant. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.}
\label{fig:ipah_venn_diagramm}
\end{figure}

>>>>

\author[1]{Devon~Kohler*}
\author[2]{Tsung-Heng~Tsai*}
\author[3]{Meena~Choi*}
\author[1]{Ting~Huang}
\author[4]{Erik~Verschueren}
\author[3]{Trent~Hinkle}
\author[1]{Olga~Vitek}
\affil[1]{Khoury College of Computer Science, Northeastern University, Boston, MA, USA}
\affil[2]{Kent State University, Kent, OH, USA}
\affil[3]{MPL, Genentech, South San Francisco, CA, USA}
\affil[4]{Galapagos, Mechelen, Antwerp, Belgium}
\affil[*]{Corresponding Authors}

@@

The scientific community widely utilizes mass spectrometry (MS)-based proteomics to quantify the abundance of proteins and their post-translational modifications (PTMs). Experiments targeting PTMs face several specific challenges which are statistical in nature. These include the low abundance of modified proteo-forms, few representative peptides that span modification sites, and convolution with abundance changes in the overall protein expression. Due to these challenges, a robust approach to estimating relative systematic changes in PTMs should combine information pertaining to PTM sites over several peptides, replicates in multiple conditions, and consider sources of confounding and variation present in the experiment. We propose a general statistical model and workflow that is both reproducible and comprehensive. The method measures modified and unmodified peptide abundance by summarizing intensities through Tukey’s median polish method. Then a model based on the family of linear mixed-effects models is fit. This model is automatically adjusted to the specific experimental design. Finally, the PTM abundances are adjusted to remove bias from changes in the overall protein. We implemented this model in the free and open-source R package $MSstatsPTM$. The approach is evaluated on computer simulations, a spike-in experiment with known ground truth, and three biological experiments with varying acquisition methods. Finally, the approach is compared against several existing methods and is shown to outperform all other methods.

@@

The signaling mechanisms that allow cells to mount a dynamic and fast response to a multitude of events are primarily facilitated by the modification of proteins at specific residues, acting as molecular on/off switches\cite{Deribe} \cite{Cohen}. Mass spectrometry-based label-free proteomics is broadly established as the tool-of-choice for unbiased and large-scale identification and quantification of proteins and their post-translational modifications (PTMs) using liquid chromatography coupled with mass spectrometry (LC-MS)\cite{Kall:2011ub} \cite{Roepstorff}. Studies targeting the post-translationally modified proteome focus either on the accurate localization of modification sites on proteins, relative or absolute quantification of a modification site's occupancy repertoire, or relative changes in occupancy across experimental conditions\cite{Mann}. Regardless of the question at hand, interrogating the modified proteome is challenging due to a number of reasons. First, the relatively lower abundance of modified proteo-forms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities, which imposes challenges in both quantification and statistical modeling. Second, contrary to the often large number of identified peptides that can be used as features to model protein abundance changes, there are relatively few representative peptides that span a modification sites, which often results in sparse, and sometimes, inherently convoluted models (i.e., single versus multiple modified sites on a single peptide). Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently convoluted with changes in the overall protein expression, making the interpretation of the results not straightforward\cite{Olsen:2013}. Therefore, a robust approach to estimate systematic relative changes in post-translational modifications, at scale, should not only combine the quantitative information pertaining to a PTM site over peptides and replicates in multiple conditions, but take into account various sources of variations and confounding factors present in the experiments.

@@

The proposed approach was implemented using the R coding language and evaluated using datasets from computer simulations, benchmark controlled mixtures, and biological investigations. The approach was compared against the commonly applied t-test and Limma methods\cite{Ritchie_15a}. The results demonstrate that by appropriately leveraging the information from the entire dataset, the proposed approach improves the reproducibility and accuracy of the estimates of PTM fold changes and results in a better calibrated type I error rate. Finally, the proposed approach is implemented as a freely available open source R package $MSstatsPTM$, available on Bioconductor, which employs similar input format as in $MSstats$ and $MSstatsTMT$, and is compatible with many acquisition methods, such as label free, DIA, DDA, and TMT\cite{Choi:2014} \cite{Huang:2020}.

@@

This section describes the datasets and experiments used to evaluate the proposed method. The datasets were chosen to provide a performance benchmark of the proposed approach against existing methods when ground truth is known, as well as represent a variety of experimental designs and acquisition methods where the approach can be applied. To benchmark the dataset, computer simulations and a spike-in experiment were used. The computer simulations vary in experimental realism; the first simulation was perfectly clean, with many replicates and no missing values, while the second simulation introduced real world qualities, such as limited modified features and missing values. The spike-in experiment took the real world qualities a step further and allowed us to compare the methods in an real experiment where ground truth is known. Finally the proposed approach was used to model three biological experiments which differed in experimental design, acquisition strategy, and use various organisms. These experiments were chosen to show the proposed method's applicability to many different types of experiments.

@@

\paragraph*{Dataset 1 : Computer simulation 1 - label free}

@@

\paragraph*{Dataset 2 : Computer simulation 2 - missing values and low replicates}

@@

\medskip

\noindent{Further details of how the simulations were generated can be found in Supplementary Sec. \ref{supp-sec:sim}. }

\subsection*{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}

@@

\subsection*{Dataset 4 : Human-1mix-TMT - ubiquitination}

The proposed approach was evaluated on an experiment where Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD)\cite{LUCHETTI2021}. Multiplex proteomics was used to quantify the abundance of total protein and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was unbalanced with two bioreplicates per condition for all conditions except for infected 1 hour. About ~95\% of the identified modified peptides derived from proteins that were quantified in the global profiling run.

Further details on the design of this experiment are provided in Supplementary Section \ref{supp-sec:ipah}.

\subsection*{Dataset 5 : Mouse-2mix-TMT - phosphorylation}

Here an experiment targeting primary murine macrophages infected with Shigella flexneri (S.flexneri)\cite{Maculins}. Multiplex proteomics was used to quantify the abundance of total protein, and phosphorylation in wild type (WT) and ATG16L1-deficient (cKO) samples, uninfected and uninfected with S.flexneri. The abundance of total protein and post-translation modifications were quantified at three time points, uninfected, early infection (45-60 minutes), and late infection (3-3.5 hours). Quantifying the total protein along with the post-translational modifications allowed us to adjust for changes in total protein and see the true impact of the site specific modifications. Two mixtures using 11-plex were ran over the six conditions. The six conditions were split between 11 channels leading to the experimental design being unbalanced. Each mixture contained two replicates per early and late WT and KO conditions. Mixture one contained one replicate of uninfected WT and two replicates of uninfected KO. Mixture two contained one replicate of uninfected KO and two uninfected WT. About ~90\% of the identified modified peptides derived from proteins that were also quantified in the global profiling run.

Further details on the design of this experiment are provided in Supplementary Sec. \ref{supp-sec:shigella}.

\subsection*{Dataset 6 : Human label free quantification - no global profiling run}

@@

Despite the important implications of PTMs in biological functions, there is a lack of general framework to summarize the available quantitative information from LC-MS data, to perform statistical inference, and to draw conclusions to characterize the quantitative properties of PTM in a statistically rigorous manner. Many investigations performed differential expression analysis of PTMs using two-sample t-test or its extensions. The approach takes as input intensities of individual features from modified peptides, or intensity ratios of modified and unmodified peptide features, and compares the mean abundance of a PTM site from one condition to another. Modifications of the t-test such as moderated t-test with Limma were also proposed\cite{Zhu}. While simple, the approach does not fully account for the sources of variations, and it is not directly applicable to experiments with complex designs, e.g., comparisons of multiple conditions, acquisition in multiple batches, etc. Additionally, while this approach can be applied to experiments targeting PTMs, there is not a self contained, straight forward implementation of the methods, making application challenging.

Isobar-PTM was developed for experiments with MS/MS quantitative strategies that employ isobaric labels such as tandem mass tags (TMT) and isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}. Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. The modeling framework, however, is not applicable for either label-free workflows or experiments with complex designs.

Further details on the specific equations and workflow of existing methods can be seen in Supplementary \ref{supp-sec:ttest}.

@@

The proposed approach takes as input a list of log-transformed intensities of spectral features, identified and quantified across LC-MS runs. The features, which are precursor ions of modified or unmodified peptides, are used to characterize the identified PTM sites and proteins. For each PTM site, the feature log-intensities of the modified peptides spanning the site are expressed using a linear mixed model in consideration of the effects of condition, run, feature and interaction between run and feature. In addition mixture may be included for features acquired via tandem mass tag (TMT) methods. The model parameters are estimated using the split-plot approach as in $MSstats$, where the feature log-intensities are first summarized into a single value per site per run in the subplot model, and the site-level summaries are then used for the inference of the PTM site abundance\cite{Choi:2014}. In the site-level summarization, Tukey's median polish (TMP), a simple and robust procedure is applied to iteratively fit a two-way additive model with the effects of run and feature, which in turn summarizes the log-intensities for each site\cite{Tukey:1977}. After summarization, the inference of the PTM site abundance in each condition is carried through fitting a model based on the family of linear mixed-effects models\cite{Bolker2009} \cite{Faraway:2006}. Statistical modeling and quantification for global proteomics data are performed by the same procedure as for PTM data.

@@

The test statistic for the hypothesis testing is the ratio of the estimate of the adjusted difference to its SE. To determine the statistical significance of the difference in terms of p-value, the test statistic is compared against the t distribution with degrees of freedom approximated by the Satterthwaite method\cite{satterthwaite:1946}. Adjustment for multiple comparisons is performed using the Benjamini-Hochberg procedure to control the false discovery rate at a desired level, e.g., $0.05$\cite{Benjamini:1995}. More details are provided in Supplementary Sec \ref{supp-sec:test}

@@

$MSstatsPTM$ gives the ability to impute missing feature intensities if desired. When values are imputed, it is assumed they are missing for reasons of low abundance. Missing features are imputed in each MS run using the Accelerated Time Failure (AFT) model\cite{Tukey:1977}. In order to impute a feature's missing values, the feature must be present in at least one MS Run. If the feature is not present across all MS Runs, the value will be left missing. Missing value imputation is done before summarization with Tukey's Median Polish in order to correct for the impact of outliers. Missing value imputation is done separately for both the PTM and global protein datasets.

@@

The statistical modeling approaches discussed above can also be extended to Tandem Mass Tag (TMT) labeling methods. TMT experiments introduce an additional source of variation in the form of different mixtures. To account for mixture variance, first feature intensities are summarized again using Tukey's Median Polish and missing values are imputed, if desired, then a new linear mixed effects model with an added term for the mixture is included. The workflow and statisical model described generally follow the methods used for modeling proteins in $MSstatsTMT$\cite{Huang:2020}. These methods are repeated to quantify both modified and unmodified peptide abundance. Once both modified and unmodified peptides are modeled, the PTM model is adjusted for changes in global protein abundance using the same methods described previously.

@@

\paragraph{Code availability}\mbox{}\\

\noindent{{\bf Bioconductor:} http://www.bioconductor.org/packages/release/bioc/html/MSstatsPTM.html}

\noindent{{\bf Github:} https://github.com/Vitek-Lab/MSstatsPTM}

@@

@@

\subsubsection*{Dataset 1 : Computer simulation 1 - label free}

The proposed approach was evaluated against two-sample t-test and Limma. 
Here we see that not accounting for changes in unmodified peptide level results in a high false discovery rate, seen in Figure \ref{fig:fdr_all_models}. When adjusting for changes in the unmodified peptide, all methods performed similarly in terms of FDR. Recall and accuracy of the proposed approach and Limma performed similarly, with Limma performing slightly better with low replicates, while t-test lagged behind. Two-sample t-test only uses data within the groups of interest while ignoring the remainder of the data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach and Limma leverage all available information, which resulted in improved power with increased number of conditions.

\subsubsection*{Dataset 2 : Computer simulation 2 - missing values and low replicates}

In this simulation, changes in unmodified peptide level still needed to be accounted for to control the FDR. Once controlling for changes in unmodified peptide, it is clear that the proposed method performed strongest across all methods, as seen in Figure \ref{fig:sim_accuracy}. The proposed method well calibrated model accuracy and the recall rate, even when the number of replicates were low. Finally, when comparing the fold change estimation across all peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change. In comparison, Limma and t-test showed much wider distributions. This fold change comparison can be seen in Figure \ref{fig:sim_boxplot}.

Further details reviewing the results of each simulation and their comparison with respect to method performance can be seen in Supplementary Section \ref{supp-sec:sim}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched}

@@

Further results of this experiment, including a comparison to t-test and fold change distribution of each method before and after adjustment, are provided in Supplementary Section \ref{supp-sec:benchmark}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Dataset 4 : Human-1mix-TMT - ubiquitination}

@@

\subsubsection*{Dataset 5 : Mouse-2mix-TMT - phosphorylation}

@@

\subsubsection*{Dataset 6 : Human label free quantification - no global profiling run}

@@

@@

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.8]{images/fig1.png}
\caption{\todo{Meena?: Can you write about how MS for PTMs is run?}}
\label{fig:workflow}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{Dataset 1 : Computer simulation 1 - label free. All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates.
}\label{fig:fdr_all_models}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{Dataset 2 : Computer simulation 2 - missing values and low replicates. The advantage of using the proposed approach is apparent when looking at the second computer simulation, which includes limited observations and the presence of missing values. In overall accuracy, the proposed method performs stronger than Limma and t-test in nearly every model. Even at lower replicates the proposed method still outperformed Limma. The lowest performaning method was $t$-test. Limma shows comparable performance to the proposed method in a clean experiment, however when real world data problems are introduced it is clear the proposed method is more robust.}
\label{fig:sim_accuracy}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/sim3_FC_boxplot.png}
\caption{Dataset 2 : Computer simulation 2 - missing values and low replicates. In the second computer simulation with missing values and low PTM features all methods correctly estimated the fold change with a median log change of .75. The proposed method in this simulation had a visibly tighter distribution around the median when compared to Limma and $t$-test. While the mean estimated fold changes were similar, the proposed method exhibited a stronger performance in correctly estimating the fold change for all peptides.}
\label{fig:sim_boxplot}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/spike_in_msstatsptm_volcano.png}
\caption{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched. Using the proposed method to model the benchmark experiment the spike in peptides (colored red) do not follow the expected log fold change before adjustment. After adjusting for changes in overall protein abundance the spike in peptides are more in line with expectation. Additionally the background grey colored peptides showed many false positives before adjustment. After adjustment the false positives were decreased considerably.
}\label{fig:spikein_prop_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/spike_in_limma_volcano.png}
\caption{Dataset 3 : SpikeIn benchmark - label-free - KGG enriched. Here the Limma method was used to model the spike in experiment. Using Limma the spike in peptides follow the expected log fold change better after adjusting for changes in protein level. However, while the fold change is much more accurate, the majority of spike in peptides do not have a significant adjusted pvalue. In terms of false positives, the results are very similar to the proposed method, with many false positives before adjustment and much fewer after.
}\label{fig:spikein_limma_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
\caption{Dataset 4 : Human-1mix-TMT - ubiquitination. The overlap of differencial modified peptides for the PTM model with and without global protein level adjustment. More PTMs became insignificant after adjustment then became significant. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein.}
\label{fig:ipah_venn_diagramm}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
\caption{Dataset 4 : Human-1mix-TMT - ubiquitination. Comparing the global profiling of protein $GSDMD\_HUMAN|P57764$ with the ubiquitination of the protein at site $K62$. The individual PSM features are shown in grey, while the feature summarization is shown in red. When looking at the summary of the modification and global protein it is clear the conditions follow different trends. Specifically, there appears to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there is a large negative change when looking at the unmodified plot. This indicates the modification is confounded with changes in the unmodified protein.}
\label{fig:ipah_profile}
\end{figure}
